---
title: "multipolygon pwsid in labeled geometries"
output: html_document
---

_Rich Pauloo_  
_Last updated `r Sys.time()`_  

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo      = TRUE, 
                      out.width = "100%", 
                      message   = FALSE,
                      error     = FALSE,
                      warning   = FALSE)
```

There are duplicate pwsid in labeled data, and these polygons are adjacent. Thus they should be joined in transformers. 

For instance:

```{r}
library(tidyverse)
library(sf)
library(fs)
library(mapview)

# mapview option for render
mapviewOptions(fgb = FALSE)

# data input location for modeling is the post-transformer staging path
staging_path <- Sys.getenv("WSB_STAGING_PATH")

# read labeled data and ignore NA pwsid
wsb_labeled <- st_read(path(staging_path, "wsb_labeled.geojson")) %>% 
  filter(!is.na(pwsid))

# multipolygon count (dc) data frame
mc <- st_drop_geometry(wsb_labeled) %>% 
  count(pwsid, sort = TRUE) %>% 
  filter(n > 1) 

mc

# multipolygon ids (mid) and data (md), remove 3 NA pwsid
mid <- mc$pwsid
md  <- filter(wsb_labeled, pwsid %in% mid) %>% 
  filter(!is.na(pwsid))

# plot dupes for visual inspection
mapview(md, zcol = "pwsid")
```


Visual inspection of duplicate pwsid indicates they are few in number (`r nrow(md)` rows and `r length(unique(md$pwsid))` unique values) and typically spatially adjacent, but in other cases, can be separated by considerable distance. This issue is addressed in the pre-modeling transformer, currently in `src/analysis/sandbox/model_explore/01_preprocess.R`. Eventually, this preprocess program will be moved to `src/predict` and the code that accomplishes the cleaning (pending consideration) may be moved to a transformer helper function. The cleaning currently implemented:  

1. unions duplicate pwsid geometries  
2. groups by pwsid and sums area  
3. recalculates radius from these areas (no convex hull, which would inflate radii for multipolygon systems with considerable space between polygons)  
4. recalculates centroids x and y (as before, these are suspect and less meaningful for non-adjacent multipolgyon systems)
